---
title: "RNA-seq Analysis of COL, log2-2, and abi5-7 Under GM & LEU Conditions"
subtitle: "Differential Expression, Clustering, and Stress-Response Profiling"
author: "Iliana Castillo-Machuca — Virginia Tech, School of Plant & Environmental Sciences"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_width: 8
    fig_height: 5
    includes:
      after_body: style.html
---

# 1. CLEAR ENVIRONMENT
```{r CLEAR-ENVIRONMENT}
#------------------------ CLEAN ENVIRONMENT ------------------------#
rm(list = ls(envir = .GlobalEnv, all.names = TRUE)) # Clear wrkspace
invisible(gc()) # Garbage collection to free memory

# Global options for consistency
options(
  max.print              = 50000, # Cap console output
  scipen                 = 0,     # Avoid sci note 999, 0 if you want sci.note
  stringsAsFactors       = FALSE, # Keep strings as character
  dplyr.summarise.inform = FALSE, # Suppress dplyr's messages
  digits                 = 4      # Number of digits to print
  )

set.seed(1234)
#------------------------ PROJECT INFO ------------------------# 
info <- list(
  Title             = "AMINO ACID TOXICITY (LOG2 ABI5) ANALYSIS",
  R_Version         = R.version$version.string,
  Working_Directory = getwd(),
  Timestamp         = format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  )

separator <- paste0("\n", strrep("=", 80), "\n")
cat("PROJECT:", info$Title, "\n")
cat(separator)

cat("R Version:", info$R_Version, "\n")
cat("Working Directory:", info$Working_Directory, "\n")
cat("Timestamp:", info$Timestamp, "\n")
cat(separator)

info$renv_active <- renv::project()
cat("renv Project:", info$renv_active, "\n")

info$Git_Commit <- system("git rev-parse --short HEAD", intern = TRUE)
cat("Git Commit:", info$Git_Commit, "\n")

#sessionInfo()
#renv::init() #first time use to make an environment with the packages
```

# 2. Packages, Load, and Installation
```{r INSTALL-PACKAGES, message=FALSE}
#------------------------ LOAD PACKAGES ------------------------#
cran_packages <- c(
  "tidyverse",    # dplyr, ggplot2, tidyr, purrr, stringr, readr, forcats
  "ggrepel",      # Add non-overlapping text labels to ggplot2
  "pheatmap",     # Clustered heat maps
  "RColorBrewer", # Color palettes
  "dendextend",   # Dendrogram
  "writexl",      # Excel export
  "ggdendro",
  "DiagrammeR",
  "gridExtra",
  "grid",
  "rmarkdown",
  "yaml",
  "BiocManager",
  "factoextra",
  "ggplot2",
  "VennDiagram",
  "grid",
  "cluster",
  "enrichplot",
  "ggVennDiagram",
  "tidytext"
)
bioc_packages <- c(
  "DESeq2",          # Differential expression analysis
  "EnhancedVolcano",
  "rstudioapi",
  "apeglm",
  "org.At.tair.db",  # Arabidopsis thaliana annotations
  "clusterProfiler"  # GO/KEGG enrichment analysis and GSEA
)

#------------------------ INSTALL CRAN PACKAGES ------------------------#
new_cran <- cran_packages[!sapply(cran_packages, requireNamespace, quietly = TRUE)]
if (length(new_cran) > 0) install.packages(new_cran)

#------------------------ INSTALL BIOCONDUCTOR ------------------------#
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

new_bioc <- bioc_packages[!sapply(bioc_packages, requireNamespace, quietly = TRUE)]
if (length(new_bioc) > 0) BiocManager::install(new_bioc, ask = FALSE)

#------------------------ LOAD PACKAGES ------------------------#
invisible(lapply(cran_packages, library, character.only = TRUE))
invisible(lapply(bioc_packages, library, character.only = TRUE))

#------------------------ WORKING DIRECTORY ------------------------# 
cat("Working directory:", getwd(), "\n")

#renv::status()
#renv::snapshot()
```
# 3. Folder Setup
```{r FOLDERS message=FALSE}
#------------------------ SET OUTPUT DIRECTORY ------------------------#  
graphics_dir <- file.path("Results/Graphics")
tables_dir   <- file.path("Results/Tables") 
procdata_dir <- file.path("Results/Processed_data")

dirs         <- c(graphics_dir, tables_dir, procdata_dir)

# Create the dirs if they don't exist
invisible( sapply(
  dirs,
    function(d) {
      if (!dir.exists(d)) {
        dir.create(d, recursive = TRUE, showWarnings = FALSE)
      } } ) )
```
# 4. Read In Raw Counts, Prep Data
```{r READ-COUNTS, message=FALSE}
#------------------------ LOAD GENOME ALIGNMENT FILES ------------------------#
featcount  <- read_tsv("./Data_Repository/Log2ABA.featcount.txt", skip = 1)
count_cols <- colnames(featcount)[7:ncol(featcount)] # backup OG names except first 7 col

read.count <- featcount %>% dplyr::select(all_of(count_cols)) %>% as.matrix()
rownames(read.count) <- featcount[[1]] #Add GENES / names

#------------------------ SET NAMES & INFO ------------------------#
info <- tibble(old = count_cols) |>
  tidyr::separate_wider_delim(
    old,
    delim    = "_",
    names    = c("rep", "genotype", "treatment"),
    too_many = "drop"
  ) |>
  mutate( genotype_clean = case_when(
      genotype == "ABI" ~ "abi5-7",
      genotype == "LOG" ~ "log2-2",
      genotype == "COL" ~ "COL",
      TRUE ~ genotype),
    new = paste(genotype_clean, treatment, rep, sep = "_"))
colnames(read.count) <- info$new

#------------------------ DESIGN SAMPLE INFO ------------------------#
sample_names <- colnames(read.count)

# Extract genotype (COL, log2-2, abi5-7)
group_setup <- stringr::str_extract(sample_names, "^(abi5-7|COL|log2-2)")
genotype    <- factor(group_setup, levels = c("COL","log2-2","abi5-7"))

# Extract treatment (GM or LEU)
treatment <- stringr::str_extract(sample_names, "(GM|LEU)")
treatment <- factor(treatment, levels = c("GM", "LEU"))

# Build sample info
sample_info <- data.frame(
  row.names = sample_names,
  genotype  = genotype,
  treatment = treatment)
stopifnot(all(rownames(sample_info) == colnames(read.count))) # Safety check

#------------------------ SAVE ------------------------#
#read.count <- tibble::rownames_to_column(as.data.frame(as.matrix(read.count)), "GeneID")
## Convert to DF with gene IDs
#raw_counts_df <- tibble::rownames_to_column(raw_counts_df, "GeneID")

#write_csv(raw_counts_df, file.path(tables_dir, "raw_counts_clean.csv"))
#writexl::write_xlsx(raw_counts_df, file.path(tables_dir, "LOG2_ABI_raw_counts_clean.xlsx"))
```
# 5. THRESHOLDS, COLORS
```{r THRESHOLDS}
#------------------------ Set Values: P.VAL, LFC ------------------------#
pad        <- 0.05
lfc_thresh <- 1.0

#------------------------ COLORS ------------------------#
colors <- c(
  "COL" = "#32BECC",   # blue curacao
  "abi5-7" = "#F9AC2F",   # citrus
  "log2-2" = "#F07BA5"    # pink cosmos
)

#------------------------ COLOR SWATCH ------------------------# 
df_colors <- data.frame(Condition = names(colors), Color = colors)

ggplot( df_colors, aes(x = Condition, y = 1, fill = Condition)) + geom_tile() + scale_fill_manual(values = colors) + theme_void() + theme( legend.position = "none", plot.margin = margin(2, 2, 2, 2)) + ggtitle("Color Swatch")
```
# 6. DESEQ2
```{r DESeq-run, message=FALSE}
#------------------------ DESEQ2 MATRIX ------------------------#
dds <- DESeqDataSetFromMatrix(
  countData = read.count,
  colData   = sample_info,
  design    = ~ genotype * treatment) 

# genotype * treatment <- asks whether the leucine response differs by genotype

#------------------------ Pre-filter ------------------------#
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

#------------------------ RUN DESEQ2 ------------------------#
dds  <- DESeq(dds)

#------------------------ CHECK & SAVE ------------------------#
#dds$treatment #check levels
#dds$genotype
resultsNames(dds)

## Save full DESeq object
#saveRDS(dds, file.path(procdata_dir, "dds_IM_H_UM.rds"))
#writeLines(resultsNames(dds), file.path(procdata_dir, "coefficient_names.txt"))
```
# 7. VST; NORM COUNTS
```{r VST}
#------------------------ VST ------------------------#
## Variance stabilizing transformation
vsd <- vst(dds, blind = FALSE) # blind ignores design
colnames(vsd) <- rownames(sample_info) ## Sample names, match to colData
stopifnot(all(colnames(vsd) == rownames(colData(dds))))  # Check
```
# 8. PCA; PLOT
```{r PCA}
#------------------------ PCA; PLOT ------------------------#
pca_data   <- plotPCA( vsd, intgroup = c("genotype", "treatment"), returnData = TRUE)
percentVar <- round( 100 * attr( pca_data, "percentVar"))

p_pca      <- ggplot(pca_data, 
                aes(PC1, PC2, color = genotype, shape = treatment)
                ) +
  geom_point(size = 5, stroke = 1.5, alpha = 0.85) +
  scale_color_manual(values = colors, name = "genotype") +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  geom_text_repel(aes(label = name),                 
                  box.padding = 0.35, 
                  max.overlaps = Inf,  # Show all labels
                  size = 3.5, 
                  segment.color = "red",
                  segment.size = 0.3,
                  bg.color = "white", 
                  bg.r = 0.1) +
  labs(title = "Sample Clustering (Variance-Stabilized Counts)",
       subtitle = paste("PC1:", round(percentVar[1],1), "% variance |",
                       "PC2:", round(percentVar[2],1), "% variance")) +
  theme_classic(base_size = 13) +
  theme(legend.position = "bottom", 
        legend.title = element_text(face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        panel.grid.minor = element_blank())
print(p_pca)

#------------------------ SAVE ------------------------#
#ggsave(file.path(graphics_dir, "LOG2_ABI5.png"), p_pca, width = 8, height = 8, dpi = 300)
```
# 9. SAMPLE DISTANCE
```{r SAMPLE DISTANCE}
#------------------------ SAMPLE DISTANCE ------------------------#
sample_dist <- dist(t(assay(vsd)))  # t samples as rows
hc <- hclust(sample_dist, method = "ward.D2")
dend_data <- ggdendro::dendro_data(hc)

#------------------------ PLOT SAMPLE DISTANCE ------------------------#
sample_dist <- dist(t(assay(vsd)))  # transpose so samples are rows
hc <- hclust(sample_dist, method = "ward.D2")
dend_data <- ggdendro::dendro_data(hc)

# Create label dataframe and add metadata
label_df <- dend_data$labels

sample_info <- as.data.frame(colData(vsd))

label_df$genotype <- sample_info$genotype[match(label_df$label, rownames(sample_info))]
label_df$treatment <- sample_info$treatment[match(label_df$label, rownames(sample_info))]
label_df$color <- colors[label_df$genotype]

# Bold LEU samples
label_df$fontface <- ifelse(label_df$treatment == "LEU", "bold", "plain")

p_dend <- ggplot() +
  geom_segment(data = ggdendro::segment(dend_data),
               aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_text(data = label_df, 
            aes(
              x = x,
              y = y,
              label = label,
              color = color,
              fontface = fontface),
            hjust = -0.1,
            size = 3.5
            ) +
  scale_color_identity() +
  coord_flip() +
  scale_y_reverse(expand = c(0.2, 0)) +
  ggdendro::theme_dendro() +
  labs(
    title = "VST Clustering (Ward.D2)",
    subtitle = "Color = Genotype, Bold = LEU"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
p_dend

#------------------------ SAVE ------------------------#
#ggsave(file.path(graphics_dir,"Distance_tree.png"), p_dend, width = 8, height = 6 )
```
# 10. SAMPLE HEATMAP
```{r}
sample_dist <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sample_dist)

rownames(sampleDistMatrix) <- paste( colData(dds)$genotype, colData(dds)$treatment,
sep = "_")
colnames(sampleDistMatrix) <- rownames(sampleDistMatrix)

p2 <- pheatmap( sampleDistMatrix, clustering_distance_rows = sample_dist, clustering_distance_cols = sample_dist, main = "Sample Distance Heatmap")
p2

#------------------------ SAVE ------------------------#
#ggsave(file.path(graphics_dir, "sample_clustering_colored.png"), 
#       p_dend, width = 12, height = 8, dpi = 300)

#------------------------ Correlation Distance ------------------------#
# Correlation distance
sampleDist_cor   <- as.dist(1 - cor(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDist_cor)

s2s <- pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDist_cor,
  clustering_distance_cols = sampleDist_cor,
  annotation_col = as.data.frame(colData(dds)[, c("genotype","treatment")]),
  main = "Sample-to-sample distance (1 - correlation)",
  treeheight_row = 75,
  treeheight_col = 60,
  cellheight = 25,
  cellwidth  = 25,
  fontsize = 9
)

```
# 11. DESEQ CONTRASTS
```{r}
#------------------------ CONTRAST FUNCTION ------------------------#
# Builds contrasts between genotype/treatment combinations.
condition_contrast <- function(dds, 
                               genotype_1, treatment_1, 
                               genotype_0, treatment_0) { 
  
  # numerator condition
  nd1 <- data.frame(genotype  = factor(genotype_1, levels  = levels(colData(dds)$genotype)),
                    treatment = factor(treatment_1, levels = levels(colData(dds)$treatment)))
  # denominator condition
  nd0 <- data.frame(genotype  = factor(genotype_0, levels  = levels(colData(dds)$genotype)),
                    treatment = factor(treatment_0, levels = levels(colData(dds)$treatment)))
  
  # Get design matrix rows for each group
  X1 <- model.matrix(design(dds), nd1)
  X0 <- model.matrix(design(dds), nd0)
  
  v  <- as.numeric(X1 - X0) # Calculate contrasts
  
  # check that they match the model coefficients
  if (length(v) != length(resultsNames(dds))) {
    stop("Model-matrix columns do not match DESeq2 coefficients.")
  } 
  v #return contrast vector
}

#------------------------ RUNNING FUNCTION ------------------------#
run_result <- function(label,
                       contrast_vector = NULL,
                       result_name     = NULL,
                       lfc_shrink      = FALSE,
                       lfc_coef        = NULL,
                       shrink_type     = "apeglm") {
# Run LFC Shrink
  if (lfc_shrink) {
    if (is.null(lfc_coef)) { # Require coefficient name for shrinkage
      stop("Provide lfc_coef when lfc_shrink = TRUE.")
    }
    # Compute shrunken results
    res <- lfcShrink( dds, coef = lfc_coef, type = shrink_type)

  } else {
    #run deseq results
    if (is.null(contrast_vector)) {
      stop("Provide contrast_vector when lfc_shrink = FALSE.")
    }
    res  <- results( dds, contrast = contrast_vector, alpha = 0.05)
  }
  res    <- res[order(res$padj, na.last = TRUE), ] # Sort by adj p val
  
  # Create data frame
  res_df <- data.frame(
    gene_id     = rownames(res),
    as.data.frame(res),
    row.names   = NULL,
    check.names = FALSE)

#  write.csv( res_df, file.path(tables_dir, paste0(label, ".csv")), row.names = FALSE)
#  writexl::write_xlsx( res_df, file.path(tables_dir, paste0(label, ".xlsx")))

  return(res) #return DESEQ results object
}
```
# 13. CONTRAST TESTING
```{r BIOLOGICAL TESTING}
#------------------------ GENOTYPE RESPONSES ------------------------#
# A. Leu response within each genotype: genes responding to amino-acid exposure
# Test which genes respond to amino-acid treatment within each genotype
res_COL_Leu <- run_result(label = "01_COL_Leu_vs_GM", contrast_vector = condition_contrast(dds, "COL", "LEU", "COL", "GM"))
res_log2_2_Leu <- run_result(label = "02_log2_2_Leu_vs_GM", contrast_vector = condition_contrast(dds, "log2-2", "LEU", "log2-2", "GM"))
res_abi5_7_Leu <- run_result(label = "03_abi5_7_Leu_vs_GM", contrast_vector = condition_contrast(dds, "abi5-7", "LEU", "abi5-7", "GM"))

#------------------------ BASELINE GENOTYPE EFFECTS ------------------------#
# B. Tests whether mutants differ from COL under control conditions
res_log2_2_vs_COL_GM <- run_result(label = "04_log2_2_vs_COL_GM", contrast_vector = condition_contrast( dds, "log2-2", "GM", "COL", "GM"))
res_abi5_7_vs_COL_GM <- run_result(label = "05_abi5_7_vs_COL_GM", contrast_vector = condition_contrast( dds, "abi5-7", "GM", "COL", "GM"))

#------------------------ GENOTYPE EFFECTS UNDER LEU ------------------------#
# C. Testing if mutants differ from COl after Leu trt
res_log2_2_vs_COL_Leu <- run_result(label = "06_log2_2_vs_COL_Leu", contrast_vector = condition_contrast( dds, "log2-2", "LEU", "COL", "LEU"))
res_abi5_7_vs_COL_Leu <- run_result(label = "07_abi5_7_vs_COL_Leu", contrast_vector = condition_contrast( dds, "abi5-7", "LEU", "COL", "LEU"))

#------------------------ GENOTYPExLEU INTERACTION ------------------------#
# D. These test whether the transcriptional response to LEU differs from COL.
res_log2_interaction <- results(dds, name = "genotypelog2.2.treatmentLEU", alpha = 0.05)
res_abi5_interaction <- results(dds, name = "genotypeabi5.7.treatmentLEU", alpha = 0.05)

#------------------------ MUTANT COMPARISON OF LEU RESPONSE ------------------------#
# E. Identifies genes responding differently to Leu in abi5-7 and log2-2
res_abi5_vs_log2_interaction <- run_result(
  label = "10_abi5_7_vs_log2_2_Leu_response",
  contrast_vector = condition_contrast(dds, "abi5-7", "LEU", "abi5-7", "GM") - 
    condition_contrast(dds, "log2-2", "LEU", "log2-2", "GM")
)

#------------------------ SUMMARY ------------------------#
count_sig <- function(res, alpha = 0.05) {
  sum(!is.na(res$padj) & res$padj < alpha)
  }

summary_table <- data.frame(
  comparison = c(
    "COL LEU vs GM",
    "log2-2 LEU vs GM",
    "abi5-7 LEU vs GM",
    "log2-2 vs COL GM",
    "abi5-7 vs COL GM",
    "log2-2 vs COL LEU",
    "abi5-7 vs COL LEU",
    "log2-2 × LEU",
    "abi5-7 × LEU"
    ),
  significant_FDR_0.05 = c(
    count_sig(res_COL_Leu),
    count_sig(res_log2_2_Leu),
    count_sig(res_abi5_7_Leu),
    count_sig(res_log2_2_vs_COL_GM),
    count_sig(res_abi5_7_vs_COL_GM),
    count_sig(res_log2_2_vs_COL_Leu),
    count_sig(res_abi5_7_vs_COL_Leu),
    count_sig(res_log2_interaction),
    count_sig(res_abi5_interaction))
  )

summary_table
```
# 14. VOLCANO
```{r VOLCANO}
#------------------------ VOLCANO ------------------------#
plot_volcano <- function(res, title){
  df <- as.data.frame(res)
  df$status <- "NS"

  df$status[
    !is.na(df$padj) &
      df$padj < 0.05 &
      df$log2FoldChange > 1 ] <- "Up"

  df$status[
    !is.na(df$padj) &
      df$padj < 0.05 &
      df$log2FoldChange < -1 ] <- "Down"

  ggplot( df, aes( log2FoldChange, -log10(pvalue)) ) +
    geom_point( aes(color = status), alpha = 0.5 ) +
    scale_color_manual(
      values = c(
        "Up" = "red",
        "Down" = "blue",
        "NS" = "grey70") ) +
    theme_classic() +
    labs(title = title)
}

#------------------------ PLOTTING & SAVE ------------------------#
v1 <- plot_volcano( res_COL_Leu, "COL: Leu vs GM")
v1
v2 <- plot_volcano( res_log2_2_Leu, "log2-2: Leu vs GM")
v2
v3 <- plot_volcano( res_abi5_7_Leu, "abi5-7: Leu vs GM")
v3

#ggsave( file.path(graphics_dir,"Figure3A_COL_Volcano.pdf"), v1, width = 7, height = 6 )
#ggsave( file.path(graphics_dir,"Figure3B_log2_2_Volcano.pdf"), v2, width = 7, height = 6 )
#ggsave( file.path(graphics_dir,"Figure3C_abi5_7_Volcano.pdf"), v3, width = 7, height = 6 )
```

# 15. SPECIFIC CANDIDATE GENES AND PLOTS
```{r CANDIDATE GENES AND PLOTS}
#------------------------ CANDIDATE GENES & PLOTS ------------------------#
candidate_genes <- c( "AT1G01010", "AT1G01020", "AT1G01030" )

for(gene in candidate_genes){
  df_gene <- plotCounts(
    dds,
    gene = gene,
    intgroup = c("genotype","treatment"),
    returnData = TRUE )
  
  p <- ggplot(df_gene,
              aes(interaction(genotype,treatment), count)) +
    geom_boxplot() +
    geom_jitter(width = 0.15)
  print(p)
}

#ggsave(file.path(graphics_dir, paste0(gene, "_expression.pdf")), p, width = 6, height = 5 ) }
```

# 16. HEATMAP ALL
```{r}
# annotations
ann_col <- as.data.frame(colData(dds)[, c("genotype", "treatment")])

plot_DE_heatmap <- function(res, expr_mat, annotation_col, title) {
  sig <- subset(
    as.data.frame(res),
    !is.na(padj) &
      padj < 0.05 &
      abs(log2FoldChange) > 1.5)

  genes <- rownames(sig)
  if(length(genes) < 2) {
    message(paste("Not enough significant genes for", title))
    return(NULL)
  }

  hm <- expr_mat[genes, ]
  # z-score
  hm <- t(scale(t(hm)))

  pheatmap(
    hm,
    annotation_col = annotation_col,
    show_rownames = FALSE,
    clustering_rows = TRUE,
    clustering_cols = TRUE,
    angle_col = 45,
        color = colorRampPalette(
      c("navy", "white", "firebrick3")
        )(100),
    main = title)
}

H1 <- plot_DE_heatmap(res_COL_Leu, mat, ann_col, "LEU vs GM")
#ggsave( file.path(graphics_dir,"01_Leu_vs_GM_heatmap.png"), H1, width = 10, height = 8)

H2 <- plot_DE_heatmap(res_log2_2_Leu, mat, ann_col, "log2-2: LEU vs GM")
#ggsave( file.path(graphics_dir,"02_log2_2_Leu_vs_GM_heatmap.png"), H2, width = 8, height = 10 )

H3 <- plot_DE_heatmap(res_abi5_7_Leu, mat, ann_col, "abi5-7: LEU vs GM")
#ggsave( file.path(graphics_dir,"03_abi5_7_Leu_vs_GM_heatmap.png"), H3, width = 8, height = 10 )
```
# 17. LEU HEATMAP
```{r LEU HEATMAP}
plot_LEU_only_heatmap <- function(res, expr_mat, annotation_col, title) {
  sig <- subset(
    as.data.frame(res),
    !is.na(padj) &
      padj < 0.05 &
      abs(log2FoldChange) > 1.5)

genes <- rownames(sig)
if(length(genes) < 2) {
  message(paste("Not enough significant genes for", title))
  return(NULL)
  }

  keep <- annotation_col$treatment == "LEU"  # Keep only LEU samples
  hm <- expr_mat[genes, keep]
  ann_sub <- annotation_col[keep, ]

  hm <- t(scale(t(hm))) # z-score
  hm <- hm[!apply(is.na(hm), 1, any), ] # Remove rows that became NaN after scaling

  pheatmap(
    hm,
    annotation_col = ann_sub,
    show_rownames = FALSE,
    clustering_rows = TRUE,
    clustering_cols = TRUE,
    angle_col = 45,
    color = colorRampPalette(
      c("navy", "white", "firebrick3")
    )(100),
    main = title)
}

H1_1 <- plot_LEU_only_heatmap(res_COL_Leu, mat, ann_col, "LEU-responsive genes (LEU samples only)")
#ggsave( file.path(graphics_dir,"01_1_Leu_vs_GM_heatmap.png"), H1_1, width = 10, height = 8)
```


```{r}
gm_genes <- rownames(
  subset(
    as.data.frame(res_abi5_7_vs_COL_GM),
    !is.na(padj) &
      padj < 0.05 &
      abs(log2FoldChange) > 1.5)
  )

leu_genes <- rownames(
  subset(
    as.data.frame(res_abi5_7_vs_COL_Leu),
    !is.na(padj) &
      padj < 0.05 &
      abs(log2FoldChange) > 1.5)
  )

cat("GM DEGs:", length(gm_genes), "\n")
cat("LEU DEGs:", length(leu_genes), "\n")
cat("Shared:", length(intersect(gm_genes, leu_genes)), "\n")
cat("GM only:", length(setdiff(gm_genes, leu_genes)), "\n")
cat("LEU only:", length(setdiff(leu_genes, gm_genes)), "\n")

abi5_int_df <- as.data.frame(res_abi5_interaction)

abi5_top <- abi5_int_df[
  order(abs(abi5_int_df$log2FoldChange),
        decreasing = TRUE),
]

head(
  abi5_top[, c(
    "log2FoldChange",
    "lfcSE",
    "stat",
    "pvalue",
    "padj"
  )],
  50)

abi5_large_effect <- subset(
  abi5_top,
  abs(log2FoldChange) > 1)

head(abi5_large_effect, 50)

keep <- ann_col$genotype %in% c("COL", "abi5-7") &
        ann_col$treatment == "LEU"

mat_sub <- mat[, keep]
ann_sub <- ann_col[keep, ]

plot_DE_heatmap(
  res_abi5_7_vs_COL_Leu,
  mat_sub,
  ann_sub,
  "abi5-7 vs COL (LEU)")

count_deg <- function(res){
  sum(
    !is.na(res$padj) &
      res$padj < 0.05 &
      abs(res$log2FoldChange) > 1.5)
  }
count_deg <- function(res){
  sum(
    !is.na(res$padj) &
      res$padj < 0.05 &
      abs(res$log2FoldChange) > 1.5)
  }

summary_table <- data.frame(
  Contrast = c(
    "COL LEU vs GM",
    "log2-2 LEU vs GM",
    "abi5-7 LEU vs GM",
    "log2-2 vs COL GM",
    "abi5-7 vs COL GM",
    "log2-2 vs COL LEU",
    "abi5-7 vs COL LEU",
    "log2-2 x LEU",
    "abi5-7 x LEU"),
  DEGs = c(
    count_deg(res_COL_Leu),
    count_deg(res_log2_2_Leu),
    count_deg(res_abi5_7_Leu),
    count_deg(res_log2_2_vs_COL_GM),
    count_deg(res_abi5_7_vs_COL_GM),
    count_deg(res_log2_2_vs_COL_Leu),
    count_deg(res_abi5_7_vs_COL_Leu),
    count_deg(res_log2_interaction),
    count_deg(res_abi5_interaction))
  )

summary_table
#write.csv(summary_table, "DEG_summary_table.csv", row.names = FALSE)
```

# 18. K MEANS CLUSTERING PLOTS
```{r}
#------------------------ KMEANS ------------------------#
# Elbow Method + Silhouette Analysis

# Significant genes
sig <- subset(as.data.frame(res_COL_Leu),
              !is.na(padj) &
                padj < 0.05 &
                abs(log2FoldChange) > 1.5)
genes <- rownames(sig)

# LEU samples only
keep <- ann_col$treatment == "LEU"
hm <- mat[genes, keep]
ann_sub <- ann_col[keep, ]

# Remove invariant genes
hm <- hm[apply(hm, 1, sd) > 0, ]

# Row z-score
hm <- t(scale(t(hm)))
hm <- hm[complete.cases(hm), ]
cat("Genes retained:", nrow(hm), "\n")

# Evaluate k values
k_range <- 2:15

wcss <- numeric(length(k_range))
silhouette_avg <- numeric(length(k_range))

set.seed(123)
for(i in seq_along(k_range)) {
  k <- k_range[i]
  km <- kmeans(hm, centers = k, nstart = 50)

  # Elbow Method
  wcss[i] <- km$tot.withinss

  # Silhouette Method
  sil <- silhouette(km$cluster, dist(hm))
  silhouette_avg[i] <- mean(sil[, 3])
}

# Elbow Plot
elbow_plot <- ggplot(data.frame(k = k_range, WCSS = wcss),
                     aes(k, WCSS)
                     ) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Elbow Method",
       x = "Number of Clusters (k)",
       y = "Within-Cluster Sum of Squares")

# Silhouette Plot
sil_plot <- ggplot(data.frame(k = k_range, silhouette = silhouette_avg),
                   aes(k, silhouette)
                   ) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Silhouette Analysis",
       x = "Number of Clusters (k)",
       y = "Average Silhouette Width")

# Display plots side-by-side
sidebyside <- grid.arrange(elbow_plot, 
             sil_plot,
             ncol = 2)
#ggsave( file.path(graphics_dir,"01_1_Kmeans.png"), sidebyside, width = 10, height = 8)

# Recommended k from silhouette
best_k <- k_range[which.max(silhouette_avg)]

cat("\nOptimal k based on silhouette score =", best_k, "\n")

# Run final k-means
km_final <- kmeans(hm, centers = best_k, nstart = 100)
table(km_final$cluster)

# K-means heatmap
H_kmeans <- pheatmap(hm, 
                     annotation_col = ann_sub, 
                     show_rownames = FALSE,
                     cluster_rows = FALSE,
                     cluster_cols = TRUE,
                     kmeans_k = best_k,
                     angle_col = 45,
                     color = colorRampPalette(
                       c("navy", "white", "firebrick3")
                       )(100),
                     main = paste0("LEU-responsive genes (k = ", best_k, ")")
                     )


#ggsave( file.path(graphics_dir,"01_1_Kmeans_k2.png"), H_kmeans, width = 10, height = 8)

# Export cluster 
 cluster_df <- data.frame(gene = rownames(hm), cluster = km_final$cluster)
 head(cluster_df)

# write.csv(
#   cluster_df,
 #  file.path(tables_dir,
#             "LEU_gene_clusters.csv"),
#   row.names = FALSE)

#  for(k in 2:8){
#  km <- kmeans(hm, centers = k,nstart = 50)
#  cat("\nK =", k, "\n")
#  print(table(km$cluster))
#  }
  
plot(k_range, silhouette_avg, type = "b", pch = 19)
```

# 19. LEU CLUSTERING HEATMAPS
```{r}
#------------------------ LEU ONLY KMEANS GENES & PLOTS ------------------------#
plot_LEU_only_heatmap <- function(res,
                                  expr_mat,
                                  annotation_col,
                                  title,
                                  padj_cutoff = 0.05,
                                  lfc_cutoff = 1.5,
                                  k = 5
                                  ) {
  # Significant genes
  sig <- as.data.frame(res)
  sig <- sig[!is.na(sig$padj) &
               sig$padj < padj_cutoff &
               abs(sig$log2FoldChange) > lfc_cutoff,
             ]
  
  if (nrow(sig) < 2){
    message("Fewer than 2 significant genes found.")
    return(NULL)
    }

  genes <- rownames(sig)
  
  # LEU samples only
  keep <- annotation_col$treatment == "LEU"
  hm <- expr_mat[genes, keep, drop = FALSE]
  ann <- annotation_col[keep, , drop = FALSE]
  
genotypes <- unique(as.character(ann$genotype))

hm_avg <- do.call(cbind,
                  lapply(genotypes,
                         function(g) {
                           idx <- ann$genotype == g
                           cat(g, "samples =", sum(idx), "\n")
                           rowMeans(hm[, idx, drop = FALSE], na.rm = TRUE)
                           }
                         ))
colnames(hm_avg) <- genotypes
rownames(hm_avg) <- rownames(hm)

  # Z-score by gene
  hm_avg <- t(scale(t(hm_avg)))

  # Remove rows containing NA/NaN
  hm_avg <- hm_avg[apply(hm_avg, 1, 
                         function(x) all(is.finite(x))), , drop = FALSE ]

  sig2 <- sig[rownames(hm_avg), ] #apply results to filtered matrix
  
# K-means clustering
  km <- kmeans(hm_avg, centers = k, nstart = 50)
  ord <- order(km$cluster)

row_ann <- data.frame(Cluster = factor(km$cluster))
rownames(row_ann) <- rownames(hm_avg)

cluster_colors <- c(
  "#1b9e77",
  "#d95f02",
  "#7570b3",
  "#e7298a",
  "#66a61e",
  "#e6ab02",
  "#a6761d",
  "#666666")

ann_colors <- list(Cluster = setNames(cluster_colors[1:k], as.character(1:k)))

ph <- pheatmap(hm_avg[ord, ],
         annotation_row = row_ann[ord, , drop = FALSE],
         annotation_colors = ann_colors,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = FALSE,
         show_colnames = TRUE,
         gaps_row = cumsum(table(km$cluster)),
         annotation_names_row = FALSE,
         main = title,
         angle_col = 45,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100))

cluster_df <- data.frame(gene = rownames(hm_avg), cluster = km$cluster, log2FC = sig2$log2FoldChange)
return(list(plot = ph,
            clusters = cluster_df,
            kmeans = km,
            cluster_sizes = table(km$cluster),
            hm_avg = hm_avg))
}

tapply(
  abs(clust_df$log2FC),
  clust_df$cluster,
  mean
)
boxplot(
  abs(log2FC) ~ cluster,
  data = clust_df
)
cluster1_genes <- subset(clust_df, cluster == 1)


#------------------------ RUN HEATMAP ------------------------#
H1_2 <- plot_LEU_only_heatmap(res = res_COL_Leu, expr_mat = mat, annotation_col = ann_col, "K-Means clustering of genotype-specific LEU-responsive genes)", k = 5)
#ggsave(file.path(graphics_dir, "01_2_cluster_leu_heatmap.png"), H1_2$plot$gtable, width = 8, height = 6, dpi = 300)

H1_2$cluster_sizes
clust_df <- H1_2$clusters # table

#------------------------ WRITE CLUSTERS ------------------------#
#write_xlsx(split(cluster_df, cluster_df$cluster),  path = file.path(tables_dir, "LEU_gene_clusters.xlsx"))
```


```{r}
# Combine simplified GO results

go_combined <- bind_rows(
  lapply(names(GO_sheets), function(cl) {

    df <- GO_sheets[[cl]]

    if (nrow(df) == 0) return(NULL)

    # Keep top 10 most significant terms
    df <- df %>%
      arrange(p.adjust) %>%
      slice_head(n = 10)

    df$Cluster <- gsub("Cluster_", "Cluster ", cl)
    # wrap GO terms
    df$Description <- str_wrap(df$Description, width = 35)
    df
  })
)

# Plot
p_go <- ggplot(go_combined,
               aes(x = Count,
                   y = reorder_within(Description,
                                      p.adjust,
                                      Cluster),
                   size = Count,
                   color = p.adjust)
               ) +
  geom_point(alpha = 0.9) +
  facet_wrap(~ Cluster,
             scales = "free_y",
             ncol = 2
             ) +
  scale_y_reordered() +
  scale_color_gradient(
    low = "hotpink",      # most significant
    high = "#233658",     # least significant
    name = "Adjusted p-value"
  ) +
  labs(title = "GO Biological Process Enrichment by K-means Cluster",
       x = "Gene Count",
       y = NULL,
       size = "Gene Count"
  ) +
  theme_bw(base_size = 12) +
  theme(strip.text = element_text(face = "bold",
                                  size = 12),
    axis.text.y = element_text(size = 12,
                               face = "bold"),
    axis.text.x = element_text(size = 11),
    legend.title = element_text(face = "bold",
                                size = 10),
    legend.text = element_text(size = 11),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14))
p_go

#ggsave(file.path(graphics_dir, "GO_clusters_faceted.png"), p_go, width = 12, height = 10, dpi = 300)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

hm_avg <- H1_2$hm_avg
km <- H1_2$kmeans

cluster_means <- hm %>%
  as.data.frame() %>%
  mutate(
    Cluster = factor(km$cluster)
  ) %>%
  group_by(Cluster) %>%
  summarise(
    across(everything(), mean),
    .groups = "drop"
  )

# Convert to long format
cluster_long <- cluster_means %>%
  pivot_longer(
    cols = -Cluster,
    names_to = "Genotype",
    values_to = "Mean_Zscore"
  )

# Plot
p_clusters <- ggplot(cluster_long,
                     aes(x = Genotype,
                         y = Mean_Zscore,
                         group = Cluster,
                         color = Cluster)
                     ) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "#1b9e77",
      "#d95f02",
      "#7570b3",
      "#e7298a",
      "#66a61e",
      "blue",
      "yellow",
      "red"
    )
  ) +
  labs(
    title = "Average Expression Profiles of K-means Clusters",
    x = NULL,
    y = "Mean Z-score"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    legend.title = element_blank()
  )
p_clusters

ggsave(
  file.path(graphics_dir, "Cluster_expression_profiles.png"),
  p_clusters,
  width = 7,
  height = 5,
  dpi = 300
)
```


```{r}
hm_avg <- H1_2$hm_avg
km <- H1_2$kmeans

plot_df <- as.data.frame(hm_avg)
plot_df$Gene <- rownames(hm_avg)
plot_df$Cluster <- factor(km$cluster)

plot_long <- plot_df %>%
  pivot_longer(cols = -c(Gene, Cluster),
               names_to = "Genotype",
               values_to = "Expression")

# Control genotype order
plot_long$Genotype <- factor(plot_long$Genotype,
                           levels = c("COL", "abi5-7", "log2-2"))
cluster_means <- plot_long %>%
  group_by(Cluster, Genotype) %>%
  summarize(Mean_Expression = mean(Expression),
            .groups = "drop")

lineclust <- ggplot() +
  geom_line(data = plot_long,
            aes(x = Genotype,
                y = Expression,
                group = Gene),
            alpha = 0.1,
            color = "grey50") +
  geom_line(data = cluster_means,
            aes(x = Genotype,
                y = Mean_Expression,
                group = 1),
            color = "black",
            linewidth = 1.2) +
  geom_point(data = cluster_means,
             aes(x = Genotype,
                 y = Mean_Expression),
             color = "hotpink",
             size = 3) +
  facet_wrap(~Cluster, ncol = 3) +
  theme_bw() +
  labs(title = "Expression Profiles of K-means Clusters",
       subtitle = "Black lines represent mean cluster profiles",
       x = NULL,
       y = "Mean Z-score") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold",
                                  size = 12),
        axis.text.x = element_text(angle = 45,
                                   face = "bold",
                                   hjust = 1),
        plot.title = element_text(face = "bold",
                                  hjust = 0.5)
        )

#ggsave(file.path(graphics_dir, "GO_clusters_line.png"), lineclust, width = 10, height = 8, dpi = 300)
```



```{r}
for(cl_name in names(GO_results)) {

  ego_res <- GO_results[[cl_name]]

  if(nrow(as.data.frame(ego_res)) == 0) next

  p <- emapplot(
    ego_res,
    showCategory = 30
  )

  ggsave(
    file.path(graphics_dir,
              paste0(cl_name,"_emap.png")),
    width = 8,
    height = 6
  )
}

p <- dotplot(
  ego_res,
  x = "Count",
  color = "p.adjust",
  showCategory = 15,
  orderBy = "p.adjust",
  font.size = 14
) +
guides(size = "none") +
theme_bw()

p <- dotplot(
  ego_res,
  x = "Count",
  color = "p.adjust",
  showCategory = 15
) +
scale_size_continuous(
  name = "Gene Count"
)

p <- dotplot(
  ego_res,
  x = "Count",
  color = "p.adjust",
  size = "Count",
  showCategory = 15
)
```


#  CLUSTERING GO ENRICHMENT ANALYSIS
```{r}
#------------------------ CLUSTERS ENRICHMENT ------------------------#
clusters <- sort(unique(clust_df$cluster))
clust_df$log2FC <- res_COL_Leu[match(clust_df$gene, rownames(res_COL_Leu)),
                               "log2FoldChange" ]

min_genes <- 5

GO_results <- list()
GO_sheets <- list()

for(cl in clusters){
  Clust_genes <- subset(clust_df, cluster == cl)$gene

  cat("Cluster", cl, "Genes:", length(Clust_genes), "\n")

  if(length(Clust_genes) >= min_genes){
    ego_res <- enrichGO(
      gene          = Clust_genes,
      OrgDb         = org.At.tair.db,
      keyType       = "TAIR",
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.05)

    # Remove redundant terms
    if(nrow(as.data.frame(ego_res)) > 0){
      ego_res <- simplify(ego_res, cutoff = 0.7, by = "p.adjust", select_fun = min)

      # Calculate term similarity matrix
      ego_res <- pairwise_termsim(ego_res)
      emapplot(ego_res, showCategory = 30)
      
      go_df <- as.data.frame(ego_res)
      cat("GO terms after simplification:", nrow(go_df), "\n" )

      GO_results[[paste0("Cluster", cl)]] <- ego_res
      GO_sheets[[paste0("Cluster_", cl)]] <- go_df
    }
  }
}

# Save all clusters in one workbook
#write_xlsx(GO_sheets, path = file.path(tables_dir, "GO_enrichment_clusters.xlsx"))

#------------------------ GO DOT PLOTS ------------------------#
go_plot_dir <- file.path(graphics_dir, "GO_dotplots")

for(cl_name in names(GO_results)) {
  ego_res <- GO_results[[cl_name]]
  
  if(nrow(as.data.frame(ego_res)) == 0) {
    cat(cl_name, ": No significant GO terms\n") 
    next
    }

p <- dotplot(ego_res, x = "Count", color = "p.adjust", size = "Count", orderBy = "p.adjust", showCategory = 15, font.size = 14) +
  ggtitle(paste("GO Enrichment -", cl_name)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

#ggsave(filename = file.path(graphics_dir, paste0(cl_name, "_GO_dotplot.png")), plot = p, width = 6, height = 6, dpi = 300)
#cat(cl_name, ": Dotplot saved\n")
}

gene_clusters <- split(cluster_df$gene, cluster_df$cluster)

cc <- compareCluster(
  geneCluster = gene_clusters,
  fun = "enrichGO",
  OrgDb = org.At.tair.db,
  keyType = "TAIR",
  ont = "BP"
)

dotplot(cc)
heatplot(cc)
```

```{r}
#------------------------ GSEA ENRICHMENT ------------------------#
gsea_res <- res_COL_Leu
gsea_res <- gsea_res[!is.na(gsea_res$log2FoldChange), ]

gene_list <- gsea_res$log2FoldChange
names(gene_list) <- rownames(gsea_res)
gene_list <- sort(gene_list, decreasing = TRUE)
head(gene_list)
gsea_BP <- gseGO(
  geneList      = gene_list,
  OrgDb         = org.At.tair.db,
  keyType       = "TAIR",
  ont           = "BP",
  minGSSize     = 10,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  verbose       = FALSE
)

gsea_df <- as.data.frame(gsea_BP)

head(gsea_df)

write_xlsx(
  gsea_df,
  file.path(
    tables_dir,
    "GSEA_BP_results.xlsx"
  )
)
gsea_BP <- pairwise_termsim(gsea_BP)
GSEA_dot <- dotplot(
  gsea_BP,
  showCategory = 20,
  split = ".sign"
) +
  facet_grid(. ~ .sign) +
  theme_bw()

ggsave(
  file.path(
    graphics_dir,
    "GSEA_dotplot.png"
  ),
  GSEA_dot,
  width = 12,
  height = 8,
  dpi = 300
)
GSEA_emap <- emapplot(
  gsea_BP,
  showCategory = 30
)

ggsave(
  file.path(
    graphics_dir,
    "GSEA_emapplot.png"
  ),
  GSEA_emap,
  width = 12,
  height = 10,
  dpi = 300
)

GSEA_ridge <- ridgeplot(
  gsea_BP,
  showCategory = 20
)

ggsave(
  file.path(
    graphics_dir,
    "GSEA_ridgeplot.png"
  ),
  GSEA_ridge,
  width = 12,
  height = 10,
  dpi = 300
)

gseaplot2(gsea_BP,
  geneSetID = 1,
  title = gsea_BP@result$Description[1]
)
gseaplot2(gsea_BP, geneSetID = "GO:0009733")
```


```{r}
#------------------------ VENNDIAGRAM ------------------------#

library(ggpubr)
library(ggVennDiagram)

sig_COL <- rownames(subset(res_COL_Leu, 
                           padj < 0.05 & 
                             abs(log2FoldChange) > 1.5))

sig_abi5 <- rownames(subset(res_abi5_7_Leu,
                            padj < 0.05 & 
                              abs(log2FoldChange) > 1.5))

venn_list <- list(COL = sig_COL, abi5 = sig_abi5)
V1 <- ggVennDiagram(venn_list, label_alpha = 0) +
  scale_fill_gradient(
    low = "white",
    high = "steelblue") + 
  theme_bw()

#ggsave(file.path(graphics_dir, "COL_vs_abi5_Venn.png"), V1, width = 8, height = 6, dpi = 300)
```


```{r}
#------------------------ DIVERGENT RESPONSES COL & LEU ------------------------#
# COMPARE COL AND abi5 LEU RESPONSES
COL <- as.data.frame(res_COL_Leu)
abi5 <- as.data.frame(res_abi5_7_Leu)

COL$gene <- rownames(COL)
abi5$gene <- rownames(abi5)

compare_df <- merge(COL,
                    abi5,
                    by = "gene", 
                    suffixes = c("_COL", "_abi5"))

compare_df <- compare_df[complete.cases(
  compare_df$log2FoldChange_COL,
  compare_df$log2FoldChange_abi5),  ]

# CORRELATION OF TRANSCRIPTIONAL RESPONSES
cor_val <- cor(compare_df$log2FoldChange_COL,
               compare_df$log2FoldChange_abi5,
               use = "complete.obs")
cat("Correlation =", round(cor_val,3), "\n")

p_cor <- ggplot(compare_df,
                aes(log2FoldChange_COL,
                    log2FoldChange_abi5)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "red") +
  stat_cor() +
  theme_bw() +
  labs(title = paste0("COL vs abi5 LEU response (r = ",
                      round(cor_val, 3), ")" ),
       x = "COL LEU vs GM log2FC",
       y = "abi5 LEU vs GM log2FC")

#ggsave(file.path(graphics_dir, "COL_vs_abi5_correlation.png"), p_cor, width = 8, height = 6)


# RESPONSE DIFFERENCE
compare_df$deltaFC <- compare_df$log2FoldChange_abi5 - compare_df$log2FoldChange_COL
compare_df <- compare_df[order(abs(compare_df$deltaFC), decreasing = TRUE ), ]

# largest differences
top_diff <- compare_df %>%
  select(gene,
         log2FoldChange_COL,
         log2FoldChange_abi5,
         deltaFC,
         padj_COL,
         padj_abi5)
#write_xlsx(list(Top_Differences = top_diff), file.path( tables_dir, "COL_vs_abi5_deltaFC.xlsx"))

# SHARED DE GENES
sig_COL <- rownames(subset(res_COL_Leu, padj < 0.05) )
sig_abi5 <- rownames(subset(res_abi5_7_Leu, padj < 0.05) )

shared_genes <- intersect(sig_COL, sig_abi5)

COL_only <- setdiff(sig_COL, sig_abi5)
abi5_only <- setdiff(sig_abi5, sig_COL)

cat("Shared:", length(shared_genes), "\n")
cat("COL only:", length(COL_only), "\n")
cat("abi5 only:", length(abi5_only), "\n")
#write_xlsx(list(Shared = data.frame(gene = shared_genes), COL_only = data.frame(gene = COL_only), abi5_only = data.frame(gene = abi5_only)), file.path(tables_dir, "Shared_and_Unique_DEGs.xlsx" ))

# GSEA
COL_rank <- res_COL_Leu$log2FoldChange
names(COL_rank) <- rownames(res_COL_Leu)

COL_rank <- COL_rank[!is.na(COL_rank)]
COL_rank <- sort(COL_rank, decreasing = TRUE)

abi5_rank <- res_abi5_7_Leu$log2FoldChange
names(abi5_rank) <- rownames(res_abi5_7_Leu)

abi5_rank <- abi5_rank[!is.na(abi5_rank)]
abi5_rank <- sort(abi5_rank, decreasing = TRUE)

gsea_COL <- gseGO(geneList = COL_rank,
                  OrgDb = org.At.tair.db,
                  keyType = "TAIR",
                  ont = "BP",
                  minGSSize = 10,
                  maxGSSize = 500,
                  pvalueCutoff = 0.05,
                  verbose = FALSE)

gsea_abi5 <- gseGO(geneList = abi5_rank,
                   OrgDb = org.At.tair.db,
                   keyType = "TAIR",
                   ont = "BP",
                   minGSSize = 10,
                   maxGSSize = 500,
                   pvalueCutoff = 0.05,
                   verbose = FALSE)

# COMPARE NES VALUES
COL_gsea <- as.data.frame(gsea_COL)
abi5_gsea <- as.data.frame(gsea_abi5)

gsea_compare <- merge(COL_gsea[,c("ID","Description","NES")],
                      abi5_gsea[,c("ID","NES")],
                      by = "ID",
                      suffixes = c("_COL","_abi5"))

gsea_compare$deltaNES <- gsea_compare$NES_abi5 - gsea_compare$NES_COL
gsea_compare <- gsea_compare[order(abs(gsea_compare$deltaNES), decreasing = TRUE ), ]

#write_xlsx(list(GSEA_NES_Comparison = gsea_compare), file.path(tables_dir, "COL_vs_abi5_GSEA.xlsx"))

# NES COMPARISON PLOT
p_nes <- ggplot(gsea_compare, aes(NES_COL, NES_abi5)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1,
              intercept = 0,
              color = "red") +
  theme_bw() +
  labs(title = "GSEA pathway comparison", x = "COL NES", y = "abi5 NES")

#ggsave(file.path(graphics_dir, "COL_vs_abi5_GSEA_NES.png"), p_nes, width = 8,  height = 6)

# CANDIDATE GENES
candidate_genes <- subset(compare_df, abs(deltaFC) > 1.5 &
                            padj_COL < 0.05 &
                            padj_abi5 < 0.05)

#write_xlsx(list(Candidate_Genes = candidate_genes), file.path(tables_dir, "COL_vs_abi5_candidate_genes.xlsx"))
cat("Candidate genes:", nrow(candidate_genes), "\n")


COL <- as.data.frame(res_COL_Leu)
abi5 <- as.data.frame(res_abi5_7_Leu)

COL$gene <- rownames(COL)
abi5$gene <- rownames(abi5)

compare_df <- merge(COL, abi5, by = "gene", suffixes = c("_COL", "_abi5"))

compare_df$deltaFC <- compare_df$log2FoldChange_abi5 - compare_df$log2FoldChange_COL

shared_sig <- subset(compare_df, 
                     padj_COL < 0.05 &
                       padj_abi5 < 0.05)

divergent_genes <- subset(shared_sig, abs(deltaFC) >= 1.5)
divergent_genes <- divergent_genes[order(abs(divergent_genes$deltaFC), decreasing = TRUE), ]

head(divergent_genes[ ,
                     c("gene",
                       "log2FoldChange_COL",
                       "log2FoldChange_abi5",
                       "deltaFC") ], 20 )

compare_df$Divergent <- abs(compare_df$deltaFC) >= 1.5 &
  compare_df$padj_COL < 0.05 &
  compare_df$padj_abi5 < 0.05

top_labels <- head(compare_df[order(abs(compare_df$deltaFC), decreasing = TRUE), ], 20 )

p <- ggplot(compare_df, 
            aes(log2FoldChange_COL, log2FoldChange_abi5)) +
  geom_point(color = "grey70", alpha = 0.4) +
  geom_point(data = subset(compare_df, Divergent), color = "red", size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_text_repel(data = top_labels, aes(label = gene), size = 3) +
  theme_bw() +
  labs(title = "COL vs abi5 LEU response",
       x = "COL log2FC",
       y = "abi5 log2FC")

#ggsave(file.path(graphics_dir, "COL_abi5_divergent_genes.png"), p, width = 8, height = 6, dpi = 300)
#write_xlsx(list( Divergent_Genes = divergent_genes ), file.path(tables_dir, "COL_abi5_divergent_genes.xlsx" ))


abi5_higher <- subset(divergent_genes, deltaFC > 1.5)$gene
COL_higher <- subset(divergent_genes, deltaFC < -1.5)$gene

ego_abi5 <- enrichGO(gene = abi5_higher,
                     OrgDb = org.At.tair.db,
                     keyType = "TAIR",
                     ont = "BP")

ego_COL <- enrichGO(gene = COL_higher,
                    OrgDb = org.At.tair.db,
                    keyType = "TAIR",
                    ont = "BP")

s_ego_abi <- simplify(ego_abi5)
s_ego_col <- simplify(ego_COL)

s_ego_abi<-pairwise_termsim(s_ego_abi)
s_ego_col<-pairwise_termsim(s_ego_col)

dotplot(s_ego_abi)
dotplot(s_ego_col)
```


```{r}
#how much overlap
col_genes <- rownames(subset(as.data.frame(res_COL_Leu),
                             padj < 0.05 &
                             abs(log2FoldChange) > 1.5))

abi_genes <- rownames(subset(as.data.frame(res_abi5_7_Leu),
                             padj < 0.05 &
                             abs(log2FoldChange) > 1.5))

length(intersect(col_genes, abi_genes))
length(setdiff(col_genes, abi_genes))
length(setdiff(abi_genes, col_genes))

gm_genes <- rownames(subset(as.data.frame(res_abi5_7_vs_COL_GM),
                            padj < 0.05 &
                            abs(log2FoldChange) > 1.5))

leu_genes <- rownames(subset(as.data.frame(res_abi5_7_vs_COL_Leu),
                             padj < 0.05 &
                             abs(log2FoldChange) > 1.5))

intersect(gm_genes, leu_genes)

head(
  as.data.frame(res_abi5_interaction)[
    order(abs(res_abi5_interaction$log2FoldChange),
          decreasing = TRUE),
  ], 50)

```

```{r}
if (params$session_info) {
  sessionInfo()
}
```